Simulating High-Dimensional Multivariate Data

### 2020-11-09 040-monte-carlo-evaluation.Rmd
2nd draft. scale up d in quad loss, drop m to 10
2nd draft. Revise MC chapter

Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency.
And, so, we'll focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately m
atching marginal and dependency parameters.

Further, we select simulaiton configurations to ultimately simulate discrete-valued RNA-seq data and the simulations will proceed in increasing complexity, leading to our motivating application in Section \@ref(examples). 
We begin wiith empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

@@

@@

*Bivariate Normal*.
Let's simulate a bivariate normal and check our correlation matching performance as N increases.
Standard bivariate normal 
Here we have BVN( $\mu_1 = \mu_2 = 0, \rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
<!-- 
Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
 -->

@@

*Bivariate Gamma*. 
Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
We vary $\rho$ across the entire possible range of correlations for each correlation type.
<!--  Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma. -->

@@

*Bivariate Negative Binomial*. 
Let's check the performance for a discrete distribution: a bivariate negative binomial.
We use values that are motivated by our RNA-seq data, namely the mean probability and sizes estimated from the data ( see [Example applications for our motivating data](examples) for estimation details): Bivariate Negative Binomial ( $prob_1 = prob_2 = `r prob`, size_1 = size_2 = 4,\rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
<!-- Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial. -->

@@

Taken together, the studies show our methology are accuracy across the entire range of possible correlation values for the rank-based dependency measures in continuous marginals.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
Table XX contains the average absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, especially near the correlation bounds.
Since the bivariate normal exactly matches all dependency measures in our scheme, we also report relative MSE of the non-normal cases to the normal simulations.

```{r ch040-combineBiSims, cache=F}
## compare relative differences in MSE for each dependency, compared to Pearson
allDat <- NULL
dat <- readRDS("./results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom') )
allDat$type <- factor(allDat$type, levels = c( 'pearson', 'kendall', 'spearman' ) )
allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
```

```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark grey, gamma in medium grey, and negative binomial in light grey. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attentuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Kendall', 'Spearman' ) 
names(typeLabs) <- c( 'pearson', 'kendall', 'spearman' )

@@

```{r ch040-biMSEtable}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMSE  <- allDat %>%
    ## filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( mae = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
knitr::kable(
  tabMSE, booktabs = TRUE,
  caption = 'Average abolute error in matching the target dependency and  '
)
```

Generally, the accuracy seems adequate for most applications, at least in these limited simulation settings.
Of course, one could choose marginals that lead to poor matching by construction.
In partice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
See [Discusion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.

## Scale up to Ultra-High Dimensions

>>>>

Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy in match dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.

@@

We select bivariate simulation configurations to ultimately simulate discrete-valued RNA-seq data and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples). 
We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
For each pair of identical margins, we vary the correlations across the entire possible range of values to evaluate the simulation's ability to obtain the theoretic bounds.
The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

@@

Table XX lists our identical-marginal, bivariate simulation configurations.
We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications for our motivating data](examples) for estimation details).
Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho$ ).

| Simulation Reps ($B$) | Correlation Types | Distribution |
|---------------------|:------------------:|--------------:|
|$\{ 1000, 10000, 100000 \}$ | $\{ \rho_P, \rho_S, \tau \}$ | $\{ { \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho ), \\ { \bf Y} \sim MVG( shape = 10, rate = 1, \rho ), \\ { \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho) \}$. |

For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).

<!-- 
*Bivariate Normal*.
Let's simulate a bivariate normal and check our correlation matching performance as N increases.
Standard bivariate normal 
Here we have BVN( $\mu_1 = \mu_2 = 0, \rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.

*Bivariate Gamma*. 
Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
We vary $\rho$ across the entire possible range of correlations for each correlation type.
Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
*Bivariate Negative Binomial*. 
Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.

 -->

@@

@@

@@

```{r ch040-combineBiSims, cache=F}
## compare relative differences in MSE for each dependency, compared to Pearson
allDat <- NULL
dat <- readRDS("./results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
```

```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attentuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )

@@

```{r ch040-biMSEtable, results='asis'}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMSE  <- allDat %>%
    ## filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( MAE = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
knitr::kable(tabMSE,
             col.names = c("No. of random vectors",
                           "Correlation type",
                           "Distribution",
                           "Mean abs. error"),
             booktabs = TRUE,
             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.'
)
```

Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
Table \@ref(tab:ch040-biMSEtable) contains the average absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.

The accuracy appears adequate for many applications.
In practice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.

## Scale up to High Dimensions
